---
title: "Bivariate Analysis"
author: "William Norfolk"
date: "11/3/2019"
output: word_document
---

This script runs bivariate/deeper analysis of the water quality dataset to produce additional/complex data visualizations. This script operates using the _processed_data_ file produced via running the _WQprocessing_ script located inside the _code_ folder and _processing_code_ subfolder loacted within this project.

Start by loading required libraries
```{r load libs}
library(readxl)
library(dplyr)
library(tidyverse)
library(forcats)
library(ggthemes)
library(plotly)
library(knitr)
library(naniar)
library(broom)
library(gridExtra)
library(zoo)
library(ggpubr)

```


Load the data and take a look at the variables.

```{r load data}

WQ_clean_data <- readRDS("../../data/processed_data/processeddata.rds")

glimpse(WQ_clean_data)

```

## Bivariate Visualization of Water Quality Parameters

Lets start by taking a look at the five water quality parameters plotted against one another with bivariate visualizations. For this data we will use jitter plots.

```{r WQ params vs each other}
temp_and_salt <- ggplot(WQ_clean_data, aes(x = salinity, y = water_temp, color = island_side)) + geom_jitter()

temp_and_salt

temp_and_ph <- ggplot(WQ_clean_data, aes(x = ph, y = water_temp, color = island_side)) + geom_jitter()

temp_and_ph

temp_and_do <- ggplot(WQ_clean_data, aes(x = dissolved_oxygen, y = water_temp, color = island_side)) + geom_jitter()

temp_and_do

temp_and_ammonia <- ggplot(WQ_clean_data, aes(x = ammonia, y = water_temp, color = island_side)) + geom_jitter()

temp_and_ammonia

salt_and_ph <- ggplot(WQ_clean_data, aes(x = salinity, y = ph, color = island_side)) + geom_jitter()

salt_and_ph

salt_and_do <- ggplot(WQ_clean_data, aes(x = salinity, y = dissolved_oxygen, color = island_side)) + geom_jitter()

salt_and_do

salt_and_ammonia <- ggplot(WQ_clean_data, aes(x = salinity, y = ammonia, color = island_side)) + geom_jitter()

salt_and_ammonia

ph_and_do <- ggplot(WQ_clean_data, aes(x = ph, y = dissolved_oxygen, color = island_side)) + geom_jitter()

ph_and_do

ph_and_ammonia <- ggplot(WQ_clean_data, aes(x = ph, y = ammonia, color = island_side)) + geom_jitter()

ph_and_ammonia

do_and_ammonia <- ggplot(WQ_clean_data, aes(x = ammonia, y = dissolved_oxygen, color = island_side)) + geom_jitter()

do_and_ammonia

```

## In Depth Seasonality

It looks like these comparisons are not very informative. This is not suprising due to the geographic spread of the site locations. While ocean waters and bay waters are quite different interms of abiotic conditions, typically the difference is primairly apparent when viewing the range of abiotic conditions rather than individual correlations. These ranges are better visualized with the violin-jitter plots produced in the _Exploratory_Data_Analysis_Location_ script. 

In our exploratory analysis we reviewed broad-scale seasonal changes (primairly using the average measure for each month). It would be worthwhile to delve deeper into the seasonal aspects of the data to see if there are any additional/stronger/weat=ker correlations we can observe. We will break down the information for each year and predictor to get a more in depth visualization of seasonality.

We will start with monthly data to view seasons.

```{r year subsets}
year_16 <- subset(WQ_clean_data, Year == "16")
year_17 <- subset(WQ_clean_data, Year == "17")
year_18 <- subset(WQ_clean_data, Year == "18")
year_19 <- subset(WQ_clean_data, Year == "19")

#To scale all figures to same axis
year_16$Month <- as.numeric(as.character(year_16$Month))
year_17$Month <- as.numeric(as.character(year_17$Month))
year_18$Month <- as.numeric(as.character(year_18$Month))
year_19$Month <- as.numeric(as.character(year_19$Month))

#To get rid of clutter in the figures
year_16 <- year_16[!is.na(year_16$island_side), ]
year_17 <- year_17[!is.na(year_17$island_side), ]
year_18 <- year_18[!is.na(year_18$island_side), ]
year_19 <- year_19[!is.na(year_19$island_side), ]
```


```{r water temp seasonality}

month_vs_temp_16 <- ggplot(year_16, aes(x = Month, y = water_temp, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(15, 40)) + geom_smooth(se = FALSE) + xlab("Month (2016)") + ylab("Water Temperature (C)") + labs(color = "Island Side")


month_vs_temp_16 

month_vs_temp_17 <- ggplot(year_17, aes(x = Month, y = water_temp, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(15, 40)) + geom_smooth(se = FALSE) + xlab("Month (2017)") + ylab("Water Temperature (C)") + labs(color = "Island Side")

month_vs_temp_17


month_vs_temp_18 <- ggplot(year_18, aes(x = Month, y = water_temp, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(15, 40)) + geom_smooth(se = FALSE) + xlab("Month (2018)") + ylab("Water Temperature (C)") + labs(color = "Island Side")

month_vs_temp_18

month_vs_temp_19 <- ggplot(year_19, aes(x = Month, y = water_temp, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(15, 40)) + geom_smooth(se = FALSE) + xlab("Month (2019)") + ylab("Water Temperature (C)") + labs(color = "Island Side")


month_vs_temp_19

temp_by_year <- ggarrange(month_vs_temp_16, month_vs_temp_17, month_vs_temp_18, month_vs_temp_19, nrow = 2, ncol = 2, common.legend = TRUE, legend = "bottom")

temp_by_year

ggsave(filename = "../../results/temp_by_year.png", plot = temp_by_year) 

```

From the looks of these figures it seems that temperature fluctuates in a manner that is to be expected (i.e. cold in the winter and warmer in the summer). It is also pretty consistent that the temperature range is greater for the bayside waters than the ocean side waters which is consistent with the previously generated violin plots from _Exploratory_Data_Analysis_Seasonal_. Aside from basic seasonality there does not appear to be any additional trends or oddities in the temperature data. 

```{r salinity seasonality}
month_vs_sal_16 <- ggplot(year_16, aes(x = Month, y = salinity, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(15, 40)) + geom_smooth(se = FALSE) + xlab("Month (2016)") + ylab("Salinity (ppt)") + labs(color = "Island Side")


month_vs_sal_16 

month_vs_sal_17 <- ggplot(year_17, aes(x = Month, y = salinity, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(15, 40)) + geom_smooth(se = FALSE) + xlab("Month (2017)") + ylab("Salinity (ppt)") + labs(color = "Island Side")

month_vs_sal_17


month_vs_sal_18 <- ggplot(year_18, aes(x = Month, y = salinity, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(15, 40)) + geom_smooth(se = FALSE) + xlab("Month (2018)") + ylab("Salinity (ppt)") + labs(color = "Island Side")

month_vs_sal_18

month_vs_sal_19 <- ggplot(year_19, aes(x = Month, y = salinity, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(15, 40)) + geom_smooth(se = FALSE) + xlab("Month (2019)") + ylab("Salinity (ppt)") + labs(color = "Island Side")

month_vs_sal_19

sal_by_year <- ggarrange(month_vs_sal_16, month_vs_sal_17, month_vs_sal_18, month_vs_sal_19, nrow = 2, ncol = 2, common.legend = TRUE, legend = "bottom")

sal_by_year

ggsave(filename = "../../results/sal_by_year.png", plot = sal_by_year) 
```
Salinity looks pretty consistent with the exploratory plots as well. Ocean side water are consistently higher in salinity than the bay side, which is expected since the bay is an estuary. Additionally, there is a wider range of salinity on the bay side, which is consistent with exploratory analysis. There is an odd jump in the salinity around April-May for 2016 and 2019. This may simply be a result of few observations taken during that time or it is possibly due to a very low rainfall year (salinitly tends to spike in the bay when rainfall is reduced). In 2017, there is a distinct salinity drop in bayside locations. The timing of this drop is correlated with the timing of Hurricane Irma landfall. We will explore more details of Irma impacts in the _Hurricane_Irma_Analysis_ script. 


```{r dissolved oxygen seasonality}
month_vs_do_16 <- ggplot(year_16, aes(x = Month, y = dissolved_oxygen, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(1, 10)) + geom_smooth(se = FALSE) + xlab("Month (2016)") + ylab("Dissolved Oxygen (mg/L)") + labs(color = "Island Side")

month_vs_do_16 

month_vs_do_17 <- ggplot(year_17, aes(x = Month, y = dissolved_oxygen, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(1, 10)) + geom_smooth(se = FALSE) + xlab("Month (2017)") + ylab("Dissolved Oxygen (mg/L)") + labs(color = "Island Side")

month_vs_do_17


month_vs_do_18 <- ggplot(year_18, aes(x = Month, y = dissolved_oxygen, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(1, 10)) + geom_smooth(se = FALSE) + xlab("Month (2018)") + ylab("Dissolved Oxygen (mg/L)") + labs(color = "Island Side")

month_vs_do_18

month_vs_do_19 <- ggplot(year_19, aes(x = Month, y = dissolved_oxygen, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(1, 10)) + geom_smooth(se = FALSE) + xlab("Month (2019)") + ylab("Dissolved Oxygen (mg/L)") + labs(color = "Island Side")

month_vs_do_19

do_by_year <- ggarrange(month_vs_do_16, month_vs_do_17, month_vs_do_18, month_vs_do_19, nrow = 2, ncol = 2, common.legend = TRUE, legend = "bottom")

do_by_year

ggsave(filename = "../../results/do_by_year.png", plot = do_by_year) 
```

Dissolved oxygen does not appear to show any distict seasonal patterns for either the bay side or the ocean side. There is a somewhat notable drop in the dissolved oxygen levels towards the end of the 2019 data for bayside samples, however or dataset ends in July of that year so it is unclear if this is just a standard fluctuation or not. It would be advised to monitor that situation if the pattern continued. 



```{r ammonia seasonality}
month_vs_amm_16 <- ggplot(year_16, aes(x = Month, y = ammonia, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(0, 0.3)) + geom_smooth(se = FALSE) + xlab("Month (2016)") + ylab("Ammonia (mg/L)") + labs(color = "Island Side")

month_vs_amm_16 

month_vs_amm_17 <- ggplot(year_17, aes(x = Month, y = ammonia, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(0, 0.3)) + geom_smooth(se = FALSE) + xlab("Month (2017)") + ylab("Ammonia (mg/L)") + labs(color = "Island Side")

month_vs_amm_17


month_vs_amm_18 <- ggplot(year_18, aes(x = Month, y = ammonia, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(0, 0.3)) + geom_smooth(se = FALSE) + xlab("Month (2018)") + ylab("Ammonia (mg/L)") + labs(color = "Island Side")

month_vs_amm_18

month_vs_amm_19 <- ggplot(year_19, aes(x = Month, y = ammonia, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(0, 0.3)) + geom_smooth(se = FALSE) + xlab("Month (2019)") + ylab("Ammonia (mg/L)") + labs(color = "Island Side")

month_vs_amm_19

amm_by_year <- ggarrange(month_vs_amm_16, month_vs_amm_17, month_vs_amm_18, month_vs_amm_19, nrow = 2, ncol = 2, common.legend = TRUE, legend = "bottom")

amm_by_year

ggsave(filename = "../../results/amm_by_year.png", plot = amm_by_year) 
```
Ammonia appears quite interesting. Most of the observations are fairly consistent with very low to negligable ammonia levels in the water. There is a distinct increase in fall 2017 which is consistent with the timing of Hurricane Irma landfall.

```{r ph seasonality}
month_vs_ph_16 <- ggplot(year_16, aes(x = Month, y = ph, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(7, 9)) + geom_smooth(se = FALSE) + xlab("Month (2016)") + ylab("pH") + labs(color = "Island Side")

month_vs_ph_16 

month_vs_ph_17 <- ggplot(year_17, aes(x = Month, y = ph, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(7, 9)) + geom_smooth(se = FALSE) + xlab("Month (2017)") + ylab("pH") + labs(color = "Island Side")

month_vs_ph_17


month_vs_ph_18 <- ggplot(year_18, aes(x = Month, y = ph, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(7, 9)) + geom_smooth(se = FALSE) + xlab("Month (2018)") + ylab("pH") + labs(color = "Island Side")

month_vs_ph_18

month_vs_ph_19 <- ggplot(year_19, aes(x = Month, y = ph, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(7, 9)) + geom_smooth(se = FALSE) + xlab("Month (2019)") + ylab("pH") + labs(color = "Island Side")

month_vs_ph_19

ph_by_year <- ggarrange(month_vs_ph_16, month_vs_ph_17, month_vs_ph_18, month_vs_ph_19, nrow = 2, ncol = 2, common.legend = TRUE, legend = "bottom")

ph_by_year

ggsave(filename = "../../results/ph_by_year.png", plot = ph_by_year) 
```
pH visualization does not show any distinct patterns across various seasons of measurement. This is consistent with expected results due to the fact that both the ocean and bay are strongly influenced by the alkaline limestone bedrock of the Keys. 

## Seasonality With Raster Plots

Next lets visualize the data using raster plots to see if we can see any differences in the correlations or additional patterns. To avoid clutter we will drop NAs from the individual predictors since they will not provide any information in the raster. 

```{r remove NA predictor values}
temp_drop_na <- WQ_clean_data[!is.na(WQ_clean_data$water_temp), ]
sal_drop_na <- WQ_clean_data[!is.na(WQ_clean_data$salinity), ]
do_drop_na <- WQ_clean_data[!is.na(WQ_clean_data$dissolved_oxygen), ]
ph_drop_na <- WQ_clean_data[!is.na(WQ_clean_data$ph), ]
amm_drop_na <- WQ_clean_data[!is.na(WQ_clean_data$ammonia), ]


```


```{r temp raster}

temp_drop_na_1 <- temp_drop_na[!is.na(temp_drop_na$island_side), ]
temp_drop_na_2 <- temp_drop_na_1[!is.na(temp_drop_na_1$site_type), ]

raster_by_site_temp <- ggplot(temp_drop_na, aes(x = Month, y = location, fill = water_temp)) + geom_raster()

raster_by_site_temp

raster_by_island_side_temp <- ggplot(temp_drop_na_1, aes(x = Month, y = island_side, fill = water_temp)) + geom_raster(interpolate = TRUE) + scale_fill_gradientn(colors = c("blue", "yellow"))

raster_by_island_side_temp

raster_by_site_type_temp <- ggplot(temp_drop_na_2, aes(x = Month, y = site_type, fill = water_temp)) + geom_raster(interpolate = TRUE) + scale_fill_gradientn(colors = c("blue", "yellow"))

raster_by_site_type_temp
```

```{r salinity raster}

sal_drop_na_1 <- sal_drop_na[!is.na(sal_drop_na$island_side), ]
sal_drop_na_2 <- sal_drop_na_1[!is.na(sal_drop_na_1$site_type), ]

raster_by_site_sal <- ggplot(sal_drop_na, aes(x = Month, y = location, fill = salinity)) + geom_raster()

raster_by_site_sal

raster_by_island_side_sal <- ggplot(sal_drop_na_1, aes(x = Month, y = island_side, fill = salinity)) + geom_raster(interpolate = TRUE) + scale_fill_gradientn(colors = c("blue", "yellow"))

raster_by_island_side_sal

raster_by_site_type_sal <- ggplot(sal_drop_na_2, aes(x = Month, y = site_type, fill = salinity)) + geom_raster(interpolate = TRUE) + scale_fill_gradientn(colors = c("blue", "yellow"))

raster_by_site_type_sal
```

```{r dissolved oxygen raster}

do_drop_na_1 <- do_drop_na[!is.na(do_drop_na$island_side), ]
do_drop_na_2 <- do_drop_na_1[!is.na(do_drop_na_1$site_type), ]

raster_by_site_do <- ggplot(do_drop_na, aes(x = Month, y = location, fill = dissolved_oxygen)) + geom_raster()

raster_by_site_do

raster_by_island_side_do <- ggplot(do_drop_na_1, aes(x = Month, y = island_side, fill = dissolved_oxygen)) + geom_raster(interpolate = TRUE) + scale_fill_gradientn(colors = c("blue", "yellow"))

raster_by_island_side_do

raster_by_site_type_do <- ggplot(do_drop_na_2, aes(x = Month, y = site_type, fill = dissolved_oxygen)) + geom_raster(interpolate = TRUE) + scale_fill_gradientn(colors = c("blue", "yellow"))

raster_by_site_type_do
```

```{r ammonia raster}

amm_drop_na_1 <- amm_drop_na[!is.na(amm_drop_na$island_side), ]
amm_drop_na_2 <- amm_drop_na_1[!is.na(amm_drop_na_1$site_type), ]

raster_by_site_amm <- ggplot(amm_drop_na, aes(x = Month, y = location, fill = ammonia)) + geom_raster()

raster_by_site_amm

raster_by_island_side_amm <- ggplot(amm_drop_na_1, aes(x = Month, y = island_side, fill = ammonia)) + geom_raster(interpolate = TRUE) + scale_fill_gradientn(colors = c("blue", "yellow"))

raster_by_island_side_amm

raster_by_site_type_amm <- ggplot(amm_drop_na_2, aes(x = Month, y = site_type, fill = ammonia)) + geom_raster(interpolate = TRUE) + scale_fill_gradientn(colors = c("blue", "yellow"))

raster_by_site_type_amm
```

```{r ph raster}

ph_drop_na_1 <- ph_drop_na[!is.na(ph_drop_na$island_side), ]
ph_drop_na_2 <- ph_drop_na_1[!is.na(ph_drop_na_1$site_type), ]

raster_by_site_ph <- ggplot(ph_drop_na, aes(x = Month, y = location, fill = ph)) + geom_raster()

raster_by_site_ph

raster_by_island_side_ph <- ggplot(ph_drop_na_1, aes(x = Month, y = island_side, fill = ph)) + geom_raster(interpolate = TRUE) + scale_fill_gradientn(colors = c("blue", "yellow"))

raster_by_island_side_ph

raster_by_site_type_ph <- ggplot(ph_drop_na_2, aes(x = Month, y = site_type, fill = ph)) + geom_raster(interpolate = TRUE) + scale_fill_gradientn(colors = c("blue", "yellow"))

raster_by_site_type_ph
```

Unsuprisingly, the raster visualizations show results consistent with the individual jitter-smooth plots above. Most of the distinct patterns are easily visualized using the raster images, however the rasters strongly emphasize the missing data values form our dataset more than any of the actual predictor information. There is a large piece of missing data in August (primairly from Seagrass/Mangrove sites) is is due to the fact that the sampling facility is typically closed for the majority of this month, and the few trips that are operated at this time tend to be reef trips since the weather is typically very good. Additionally, there is a data gap for Patch Reef/Hardbottom locations in January. Is likely due to the fact that winter winds typicaly make patch/hardbottom site more difficult to access safely at this time. 

Though visualy appealing, the line graphs produced above show data patterns that are more consistent with the actual measures of interest thus rasters will be relegated to supplemental information. 


## Most Frequent Site Visited By Type

Lastly in our bivariate analysis we will zoom into three specific locations that have the most recorded observations of Seagrass/Mangrove, Patch Reef/Hardbottom, and Coral Reef site types to visualize specific seasonal impacts on these locations. 

First we must determine the most abundant locations of each type and subset the data. 
```{r}
patch_hb_subset <- subset(WQ_clean_data, site_type == "Patch Reef/Hardbottom")
sg_mg_subset <- subset(WQ_clean_data, site_type == "Seagrass/Mangrove")
coral_reef_subset <- subset(WQ_clean_data, site_type == "Coral Reef")

patch_plot <- ggplot(patch_hb_subset, aes(x = location)) + geom_bar() + coord_flip()
sg_mg_plot <- ggplot(sg_mg_subset, aes(x = location)) + geom_bar() + coord_flip()
coral_plot <- ggplot(coral_reef_subset, aes(x = location)) + geom_bar() + coord_flip()

grid.arrange(patch_plot, sg_mg_plot, coral_plot, nrow = 1)


```

So it looks like our most numerous observation for Patch Reef/Hardbottom is Rodriguez Key, Seagrass/Mangrove is Tarpon Basin, and Coral Reef is Grecian Dry Rocks. This is unsuprising as all of these sites are in close proximity to the marina and are typically better sheltered on windy days, thus permitting more frequent access. 

Let's subset these locations and visualize.

```{r}

roddy_subset <- subset(WQ_clean_data, location == "Rodriguez Key")
tarpon_subset <- subset(WQ_clean_data, location == "Tarpon Basin")
grecian_subset <- subset(WQ_clean_data, location == "Grecian Dry Rocks")

group_subset <- rbind(roddy_subset, tarpon_subset, grecian_subset)
```

We will start by grouping them all togeather with each of the five predictors and then compare seasonal fluctuations across the entire sampling timeframe. 

```{r}
freq_site_temp <- ggplot(group_subset, aes(x = Month, y = water_temp, color = location)) + geom_jitter()

freq_site_temp

freq_site_sal <- ggplot(group_subset, aes(x = Month, y = salinity, color = location)) + geom_jitter()

freq_site_sal

freq_site_amm <- ggplot(group_subset, aes(x = Month, y = ammonia, color = location)) + geom_jitter()

freq_site_sal

freq_site_do <- ggplot(group_subset, aes(x = Month, y = dissolved_oxygen, color = location)) + geom_jitter()

freq_site_do

freq_site_ph <- ggplot(group_subset, aes(x = Month, y = ph, color = location)) + geom_jitter()

freq_site_ph
```

These plots are very consistent with the plots of the full data set.


Lastly lets take the seasonal means for each parameter and apply them to our three sites of interest. We will generate line graphs to show specific fluctuation between sites.

```{r}
mean_temp_roddy <- aggregate(water_temp ~ Month + Year, roddy_subset, mean) 

mean_temp_roddy$Date <- as.yearmon(paste(mean_temp_roddy$Month, mean_temp_roddy$Year, sep = "."), format = "%m.%Y")

mean_temp_plot_roddy <- mean_temp_roddy %>% ggplot(aes(x = Date, y = water_temp)) + geom_line() + xlab("Year") + ylab("Water Temp (C)") + ggtitle("Mean Water Temp Rodriguez Key") + scale_y_continuous(limits=c(18, 35))

mean_temp_plot_roddy

mean_temp_tarpon <- aggregate(water_temp ~ Month + Year, tarpon_subset, mean) 

mean_temp_tarpon$Date <- as.yearmon(paste(mean_temp_tarpon$Month, mean_temp_tarpon$Year, sep = "."), format = "%m.%Y")

mean_temp_plot_tarpon <- mean_temp_tarpon %>% ggplot(aes(x = Date, y = water_temp)) + geom_line() + xlab("Year") + ylab("Water Temp (C)") + ggtitle("Mean Water Temp Tarpon Basin") + scale_y_continuous(limits=c(18, 35))

mean_temp_plot_tarpon

mean_temp_grecian <- aggregate(water_temp ~ Month + Year, grecian_subset, mean) 

mean_temp_grecian$Date <- as.yearmon(paste(mean_temp_grecian$Month, mean_temp_grecian$Year, sep = "."), format = "%m.%Y")

mean_temp_plot_grecian <- mean_temp_grecian %>% ggplot(aes(x = Date, y = water_temp)) + geom_line() + xlab("Year") + ylab("Water Temp (C)") + ggtitle("Mean Water Temp Grecian Dry Rocks") + scale_y_continuous(limits=c(18, 35))

mean_temp_plot_grecian

grid.arrange(mean_temp_plot_roddy, mean_temp_plot_tarpon, mean_temp_plot_grecian, nrow = 3)


```

Temperature looks pretty seasonally consistent withe the other plots previously generated.

```{r}
mean_sal_roddy <- aggregate(salinity ~ Month + Year, roddy_subset, mean) 

mean_sal_roddy$Date <- as.yearmon(paste(mean_sal_roddy$Month, mean_sal_roddy$Year, sep = "."), format = "%m.%Y")

mean_sal_plot_roddy <- mean_sal_roddy %>% ggplot(aes(x = Date, y = salinity)) + geom_line() + xlab("Year") + ylab("Salinity (ppt)") + ggtitle("Mean Salinity Rodriguez Key") + scale_y_continuous(limits=c(20, 40))

mean_sal_plot_roddy

mean_sal_tarpon <- aggregate(salinity ~ Month + Year, tarpon_subset, mean) 

mean_sal_tarpon$Date <- as.yearmon(paste(mean_sal_tarpon$Month, mean_sal_tarpon$Year, sep = "."), format = "%m.%Y")

mean_sal_plot_tarpon <- mean_sal_tarpon %>% ggplot(aes(x = Date, y = salinity)) + geom_line() + xlab("Year") + ylab("Salinity (ppt)") + ggtitle("Mean Salinity Tarpon Basin") + scale_y_continuous(limits=c(20, 40))

mean_sal_plot_tarpon

mean_sal_grecian <- aggregate(salinity ~ Month + Year, grecian_subset, mean) 

mean_sal_grecian$Date <- as.yearmon(paste(mean_sal_grecian$Month, mean_sal_grecian$Year, sep = "."), format = "%m.%Y")

mean_sal_plot_grecian <- mean_sal_grecian %>% ggplot(aes(x = Date, y = salinity)) + geom_line() + xlab("Year") + ylab("Salinity (ppt)") + ggtitle("Mean Salinity Grecian") + scale_y_continuous(limits=c(20, 40))

mean_sal_plot_grecian

grid.arrange(mean_sal_plot_roddy, mean_sal_plot_tarpon, mean_sal_plot_grecian, nrow = 3)
```

Salinity shows the distinct between ocean and bayside sites as expected. Additionall we can see a dip in the salinity of Tarpon Basin sharply near the Irma timeframe, this suggests that the bay may have been more impacted by the storm. This makes sense as the bay is shallower. There is also an interesting decline in the salinity of Grecian Dry Rocks at the end of 2019, I am unsure as to what may be the cause of this decline. 

```{r}
mean_do_roddy <- aggregate(dissolved_oxygen ~ Month + Year, roddy_subset, mean) 

mean_do_roddy$Date <- as.yearmon(paste(mean_do_roddy$Month, mean_do_roddy$Year, sep = "."), format = "%m.%Y")

mean_do_plot_roddy <- mean_do_roddy %>% ggplot(aes(x = Date, y = dissolved_oxygen)) + geom_line() + xlab("Year") + ylab("Dissolved Oxygen (mg/L)") + ggtitle("Mean Dissolved Oxygen Rodriguez Key") + scale_y_continuous(limits=c(2, 10))

mean_do_plot_roddy

mean_do_tarpon <- aggregate(dissolved_oxygen ~ Month + Year, tarpon_subset, mean) 

mean_do_tarpon$Date <- as.yearmon(paste(mean_do_tarpon$Month, mean_do_tarpon$Year, sep = "."), format = "%m.%Y")

mean_do_plot_tarpon <- mean_do_tarpon %>% ggplot(aes(x = Date, y = dissolved_oxygen)) + geom_line() + xlab("Year") + ylab("Dissolved Oxygen (mg/L)") + ggtitle("Mean Dissolved Oxygen Tarpon Basin") + scale_y_continuous(limits=c(2, 10))

mean_do_plot_tarpon

mean_do_grecian <- aggregate(dissolved_oxygen ~ Month + Year, grecian_subset, mean) 

mean_do_grecian$Date <- as.yearmon(paste(mean_do_grecian$Month, mean_do_grecian$Year, sep = "."), format = "%m.%Y")

mean_do_plot_grecian <- mean_do_grecian %>% ggplot(aes(x = Date, y = dissolved_oxygen)) + geom_line() + xlab("Year") + ylab("Dissolved Oxygen (mg/L)") + ggtitle("Mean Dissolved Oxygen Grecian") + scale_y_continuous(limits=c(2, 10))

mean_do_plot_grecian

grid.arrange(mean_do_plot_roddy, mean_do_plot_tarpon, mean_do_plot_grecian, nrow = 3)



```

Dissolved oxygen is rather interesting. The general patterns of the graphs are similar to the expected variation between sites, however there is a notable decrease in Tarpon Basin around the Irma timeline. This also suggests the bay was strongly impacted by the storm. There is also an interesting spike in DO in early 2016 for Grecian Dry Rocks. I am unsure as to the cause of this.

```{r}
mean_amm_roddy <- aggregate(ammonia ~ Month + Year, roddy_subset, mean) 

mean_amm_roddy$Date <- as.yearmon(paste(mean_amm_roddy$Month, mean_amm_roddy$Year, sep = "."), format = "%m.%Y")

mean_amm_plot_roddy <- mean_amm_roddy %>% ggplot(aes(x = Date, y = ammonia)) + geom_line() + xlab("Year") + ylab("Ammonia (mg/L)") + ggtitle("Mean Ammonia Rodriguez Key") + scale_y_continuous(limits=c(0, 1))

mean_amm_plot_roddy

mean_amm_tarpon <- aggregate(ammonia ~ Month + Year, tarpon_subset, mean) 

mean_amm_tarpon$Date <- as.yearmon(paste(mean_amm_tarpon$Month, mean_amm_tarpon$Year, sep = "."), format = "%m.%Y")

mean_amm_plot_tarpon <- mean_amm_tarpon %>% ggplot(aes(x = Date, y = ammonia)) + geom_line() + xlab("Year") + ylab("Ammonia (mg/L)") + ggtitle("Mean Ammonia Tarpon Basin") + scale_y_continuous(limits=c(0, 1))

mean_amm_plot_tarpon

mean_amm_grecian <- aggregate(ammonia ~ Month + Year, grecian_subset, mean) 

mean_amm_grecian$Date <- as.yearmon(paste(mean_amm_grecian$Month, mean_amm_grecian$Year, sep = "."), format = "%m.%Y")

mean_amm_plot_grecian <- mean_amm_grecian %>% ggplot(aes(x = Date, y = ammonia)) + geom_line() + xlab("Year") + ylab("Ammonia (mg/L)") + ggtitle("Mean Ammonia Grecian") + scale_y_continuous(limits=c(0, 1))

mean_amm_plot_grecian

grid.arrange(mean_amm_plot_roddy, mean_amm_plot_tarpon, mean_amm_plot_grecian, nrow = 3)
```

Ammonia is similar to other visuals created above, relatively flat across the board with intermittant spikes and a substantial spike around Irma.

```{r}
mean_ph_roddy <- aggregate(ph ~ Month + Year, roddy_subset, mean) 

mean_ph_roddy$Date <- as.yearmon(paste(mean_ph_roddy$Month, mean_ph_roddy$Year, sep = "."), format = "%m.%Y")

mean_ph_plot_roddy <- mean_ph_roddy %>% ggplot(aes(x = Date, y = ph)) + geom_line() + xlab("Year") + ylab("pH") + ggtitle("Mean pH Rodriguez Key") + scale_y_continuous(limits=c(7, 9))

mean_ph_plot_roddy

mean_ph_tarpon <- aggregate(ph ~ Month + Year, tarpon_subset, mean) 

mean_ph_tarpon$Date <- as.yearmon(paste(mean_ph_tarpon$Month, mean_ph_tarpon$Year, sep = "."), format = "%m.%Y")

mean_ph_plot_tarpon <- mean_ph_tarpon %>% ggplot(aes(x = Date, y = ph)) + geom_line() + xlab("Year") + ylab("pH") + ggtitle("Mean pH Tarpon Basin") + scale_y_continuous(limits=c(7, 9))

mean_ph_plot_tarpon

mean_ph_grecian <- aggregate(ph ~ Month + Year, grecian_subset, mean) 

mean_ph_grecian$Date <- as.yearmon(paste(mean_ph_grecian$Month, mean_ph_grecian$Year, sep = "."), format = "%m.%Y")

mean_ph_plot_grecian <- mean_ph_grecian %>% ggplot(aes(x = Date, y = ph)) + geom_line() + xlab("Year") + ylab("pH") + ggtitle("Mean pH Grecian") + scale_y_continuous(limits=c(7, 9))

mean_ph_plot_grecian

grid.arrange(mean_ph_plot_roddy, mean_ph_plot_tarpon, mean_ph_plot_grecian, nrow = 3)
```
 pH is also quite interesting. Similar to dissolved oxygen, there is a notable drop in pH around hurricane Irma land fall that is not present in the group data. This again suggests the bay was strongly affected by this storm.




## Conclusions

Overall, the there seems to be little seasonal fluctuation in ammonia, dissolved oxygen and pH measurements. This is not suprising due to the fact that these parameters are largely independent of one another (barring some unique event). Water temperature shows standard seasonal fluctuation with warmer and cooler months, and bays side waters exhibit a wider ranger of temperatures. Salinity is reasonable unique in terms of observations. Oceanside water have distinctly greater salinity for the vast majority of the year and have a very narrow range. Bayside waters typically have a reduced salinity, however large spikes upward can be associated with the dry season. 

Despite these overall trends, there is a notable variation in some of the parameters consistent with the timing of Hurricane Irma landfall (September 2017). Ammonia most noteably spkies drastically at this time. This is likely due to hurricane-induced nutrint influx. Salinity drops substantially from the normal levels, likely due to storm water influx. Water temperature shows a notable decrease, but it is likely this can be attributed to standard seasonal fluctuation. Both pH and dissolved oxygen do not show any unusual trends with the group data, but show notable changes when aplied to Tarpon Basin measures. Hurricane Irma warrents further investigation and will be analyzed in the _hurricane_irma_analysis_ script. It should be noted that post Hirricane Irma observations are limited in number due to the inability to collect samples directly following the storm. 






