---
title: "Bivariate Analysis"
author: "William Norfolk"
date: "11/3/2019"
output: word_document
---

This script runs bivariate/deeper analysis of the water quality dataset to produce additional/complex data visualizations. This script operates using the _processed_data_ file produced via running the _WQprocessing_ script located inside the _code_ folder and _processing_code_ subfolder loacted within this project.

Start by loading required libraries
```{r load libs}
library(readxl)
library(dplyr)
library(tidyverse)
library(forcats)
library(ggthemes)
library(plotly)
library(knitr)
library(naniar)
library(broom)
library(gridExtra)
library(zoo)
library(ggpubr)

```


Load the data and take a look at the variables.

```{r load data}

WQ_clean_data <- readRDS("../../data/processed_data/processeddata.rds")

glimpse(WQ_clean_data)

```

Lets start by taking a look at the five water quality parameters plotted against one another with bivariate visualizations. For this data we will use jitter plots.

```{r WQ params vs each other}
temp_and_salt <- ggplot(WQ_clean_data, aes(x = salinity, y = water_temp, color = island_side)) + geom_jitter()

temp_and_salt

temp_and_ph <- ggplot(WQ_clean_data, aes(x = ph, y = water_temp, color = island_side)) + geom_jitter()

temp_and_ph

temp_and_do <- ggplot(WQ_clean_data, aes(x = dissolved_oxygen, y = water_temp, color = island_side)) + geom_jitter()

temp_and_do

temp_and_ammonia <- ggplot(WQ_clean_data, aes(x = ammonia, y = water_temp, color = island_side)) + geom_jitter()

temp_and_ammonia

salt_and_ph <- ggplot(WQ_clean_data, aes(x = salinity, y = ph, color = island_side)) + geom_jitter()

salt_and_ph

salt_and_do <- ggplot(WQ_clean_data, aes(x = salinity, y = dissolved_oxygen, color = island_side)) + geom_jitter()

salt_and_do

salt_and_ammonia <- ggplot(WQ_clean_data, aes(x = salinity, y = ammonia, color = island_side)) + geom_jitter()

salt_and_ammonia

ph_and_do <- ggplot(WQ_clean_data, aes(x = ph, y = dissolved_oxygen, color = island_side)) + geom_jitter()

ph_and_do

ph_and_ammonia <- ggplot(WQ_clean_data, aes(x = ph, y = ammonia, color = island_side)) + geom_jitter()

ph_and_ammonia

do_and_ammonia <- ggplot(WQ_clean_data, aes(x = ammonia, y = dissolved_oxygen, color = island_side)) + geom_jitter()

do_and_ammonia

```

It looks like these comparisons are not terribly informative. This is not suprising due to the geographic spread of the site locations. While ocean waters and bay waters are quite different interms of abiotic conditions, typically the difference is primairly apparent when viewing the range of abiotic conditions rather than individual correlations. These ranges are better visualized with the violin-jitter plots produced in the _Exploratory_Data_Analysis_Location_ script. 

It may be more informative to view seasonal and annual changes in conditions using bivariate analysis. We will break down the information for each year and predictor to get a more in depth visualization of seasonality.

We will start with monthly data to view seasons.

```{r year subsets}
year_16 <- subset(WQ_clean_data, Year == "16")
year_17 <- subset(WQ_clean_data, Year == "17")
year_18 <- subset(WQ_clean_data, Year == "18")
year_19 <- subset(WQ_clean_data, Year == "19")

#To scale all figures to same axis
year_16$Month <- as.numeric(as.character(year_16$Month))
year_17$Month <- as.numeric(as.character(year_17$Month))
year_18$Month <- as.numeric(as.character(year_18$Month))
year_19$Month <- as.numeric(as.character(year_19$Month))

#To get rid of clutter in the figures
year_16 <- year_16[!is.na(year_16$island_side), ]
year_17 <- year_17[!is.na(year_17$island_side), ]
year_18 <- year_18[!is.na(year_18$island_side), ]
year_19 <- year_19[!is.na(year_19$island_side), ]
```


```{r water temp seasonality}

month_vs_temp_16 <- ggplot(year_16, aes(x = Month, y = water_temp, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(15, 40)) + geom_smooth(se = FALSE) + xlab("Month (2016)") + ylab("Water Temperature (C)")

month_vs_temp_16 

month_vs_temp_17 <- ggplot(year_17, aes(x = Month, y = water_temp, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(15, 40)) + geom_smooth(se = FALSE) + xlab("Month (2017)") + ylab("Water Temperature (C)")

month_vs_temp_17


month_vs_temp_18 <- ggplot(year_18, aes(x = Month, y = water_temp, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(15, 40)) + geom_smooth(se = FALSE) + xlab("Month (2018)") + ylab("Water Temperature (C)")

month_vs_temp_18

month_vs_temp_19 <- ggplot(year_19, aes(x = Month, y = water_temp, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(15, 40)) + geom_smooth(se = FALSE) + xlab("Month (2019)") + ylab("Water Temperature (C)")

month_vs_temp_19

ggarrange(month_vs_temp_16, month_vs_temp_17, month_vs_temp_18, month_vs_temp_19, nrow = 2, ncol = 2, common.legend = TRUE, legend = "bottom")

```

From the looks of these figures it seems that temperature fluctuates in a manner that is to be expected (i.e. cold in the winter and warmer in the summer). It is also pretty consistent that the temperature range is greater for the bayside waters than the ocean side waters which is consistent withe previously generated violin plots from _Exploratory_Data_Analysis_Seasonal_.

```{r}
month_vs_sal_16 <- ggplot(year_16, aes(x = Month, y = salinity, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(15, 40)) + geom_smooth(se = FALSE) + xlab("Month (2016)") + ylab("Salinity (ppt)")

month_vs_sal_16 

month_vs_sal_17 <- ggplot(year_17, aes(x = Month, y = salinity, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(15, 40)) + geom_smooth(se = FALSE) + xlab("Month (2017)") + ylab("Salinity (ppt)")

month_vs_sal_17


month_vs_sal_18 <- ggplot(year_18, aes(x = Month, y = salinity, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(15, 40)) + geom_smooth(se = FALSE) + xlab("Month (2018)") + ylab("Salinity (ppt)")

month_vs_sal_18

month_vs_sal_19 <- ggplot(year_19, aes(x = Month, y = salinity, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(15, 40)) + geom_smooth(se = FALSE) + xlab("Month (2019)") + ylab("Salinity (ppt)")

month_vs_sal_19

ggarrange(month_vs_sal_16, month_vs_sal_17, month_vs_sal_18, month_vs_sal_19, nrow = 2, ncol = 2, common.legend = TRUE, legend = "bottom")
```
Salinity loos pretty consistent with the exploratory plots as well. Ocean side water are consistently higher in salinity than the bay side, which is expected since the bay is an estuary. Additionally, there is wa wider range of salinity on the bay side, this is consistent with exploratory analysis. There is an odd jump in the salinity around April. This may simply be a result of few observations taken during that time or it is possibly due to a very low rainfall year (salinitly tends to spike in the bay when rainfall is reduced). We will explore this further in discussion.


```{r}
month_vs_do_16 <- ggplot(year_16, aes(x = Month, y = dissolved_oxygen, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(1, 10)) + geom_smooth(se = FALSE) + xlab("Month (2016)") + ylab("Dissolved Oxygen (mg/L")

month_vs_do_16 

month_vs_do_17 <- ggplot(year_17, aes(x = Month, y = dissolved_oxygen, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(1, 10)) + geom_smooth(se = FALSE) + xlab("Month (2017)") + ylab("Dissolved Oxygen (mg/L")

month_vs_do_17


month_vs_do_18 <- ggplot(year_18, aes(x = Month, y = dissolved_oxygen, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(1, 10)) + geom_smooth(se = FALSE) + xlab("Month (2018)") + ylab("Dissolved Oxygen (mg/L")

month_vs_do_18

month_vs_do_19 <- ggplot(year_19, aes(x = Month, y = dissolved_oxygen, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(1, 10)) + geom_smooth(se = FALSE) + xlab("Month (2019)") + ylab("Dissolved Oxygen (mg/L")

month_vs_do_19

ggarrange(month_vs_do_16, month_vs_do_17, month_vs_do_18, month_vs_do_19, nrow = 2, ncol = 2, common.legend = TRUE, legend = "bottom")
```

Dissolved oxygen does not appear to show any distict patterns across all years sampled. Though the 2019 plot does show clearly a large upswing in total number of observations taken. 



```{r}
month_vs_amm_16 <- ggplot(year_16, aes(x = Month, y = ammonia, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(0, 0.3)) + geom_smooth(se = FALSE) + xlab("Month (2016)") + ylab("Ammonia (mg/L)")

month_vs_amm_16 

month_vs_amm_17 <- ggplot(year_17, aes(x = Month, y = ammonia, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(0, 0.3)) + geom_smooth(se = FALSE) + xlab("Month (2017)") + ylab("Ammonia (mg/L)")

month_vs_amm_17


month_vs_amm_18 <- ggplot(year_18, aes(x = Month, y = ammonia, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(0, 0.3)) + geom_smooth(se = FALSE) + xlab("Month (2018)") + ylab("Ammonia (mg/L)")

month_vs_amm_18

month_vs_amm_19 <- ggplot(year_19, aes(x = Month, y = ammonia, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(0, 0.3)) + geom_smooth(se = FALSE) + xlab("Month (2019)") + ylab("Ammonia (mg/L)")

month_vs_amm_19

ggarrange(month_vs_amm_16, month_vs_amm_17, month_vs_amm_18, month_vs_amm_19, nrow = 2, ncol = 2, common.legend = TRUE, legend = "bottom")
```
Ammonia appears quite interesting. Most of the observations are fairly consistent with very low to negligable ammonia levels in the water. There is a distinct increase in fall 2017 which is consistent with the timing of Hurricane Irma landfall.

```{r}
month_vs_ph_16 <- ggplot(year_16, aes(x = Month, y = ph, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(7, 9)) + geom_smooth(se = FALSE) + xlab("Month (2016)") + ylab("pH")

month_vs_ph_16 

month_vs_ph_17 <- ggplot(year_17, aes(x = Month, y = ph, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(7, 9)) + geom_smooth(se = FALSE) + xlab("Month (2017)") + ylab("pH")

month_vs_ph_17


month_vs_ph_18 <- ggplot(year_18, aes(x = Month, y = ph, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(7, 9)) + geom_smooth(se = FALSE) + xlab("Month (2018)") + ylab("pH")

month_vs_ph_18

month_vs_ph_19 <- ggplot(year_19, aes(x = Month, y = ph, color = island_side)) + geom_jitter(alpha = 0.25) + scale_x_continuous(limits=c(1, 12), breaks = 1:12) + scale_y_continuous(limits=c(7, 9)) + geom_smooth(se = FALSE) + xlab("Month (2019)") + ylab("pH")

month_vs_ph_19

ggarrange(month_vs_ph_16, month_vs_ph_17, month_vs_ph_18, month_vs_ph_19, nrow = 2, ncol = 2, common.legend = TRUE, legend = "bottom")
```
pH visualization does not show any distinct patterns across various seasons of measurement. This is consistent with expected results due to the fact that both the ocean and bay are strongly influenced by the alkaline limestone bedrock of the Keys. 

Month vs params

Year vs params

raster with site names 1 pred and and island side











