---
title: "Exploratory Data Analysis"
author: "William Norfolk"
date: "10/9/2019"
output: word_document
---

Load the required libraries for exploratory analysis.

```{r}
library(readxl)
library(dplyr)
library(tidyverse)
library(forcats)
library(ggthemes)
library(plotly)
library(knitr)
library(naniar)
library(broom)
```

Load the processed data from the RDS. Then take a look!

```{r}
WQ_clean_data <- readRDS("../../data/processed_data/processeddata.rds")

glimpse(WQ_clean_data)
```


Now lets generate some plots to compare ocean vs bay. 

Looks like pH is pretty consistent between the ocean and bayside.

```{r}
WQ_clean_data %>% ggplot() + 
  geom_boxplot(aes(x = island_side, y = ph), na.rm = TRUE)

WQ_clean_data %>% ggplot() + 
  geom_jitter(aes(x = island_side, y = ph))
```

Looks like salinity may be a little more interesting. Ocean side appears to have a narrower range of values compared to the bayside,  which is expected due to the size of the body of water.

```{r}

WQ_clean_data %>% ggplot() + 
  geom_boxplot(aes(x = island_side, y = salinity), na.rm = TRUE)

WQ_clean_data %>% ggplot() + 
  geom_jitter(aes(x = island_side, y = salinity))
```



```{r}
WQ_clean_data %>% ggplot() + 
  geom_boxplot(aes(x = island_side, y = water_temp), na.rm = TRUE)

WQ_clean_data %>% ggplot() + 
  geom_jitter(aes(x = island_side, y = water_temp))
```

```{r}
WQ_clean_data %>% ggplot() + 
  geom_boxplot(aes(x = island_side, y = dissolved_oxygen), na.rm = TRUE)

WQ_clean_data %>% ggplot() + 
  geom_jitter(aes(x = island_side, y = dissolved_oxygen))
```

```{r}
WQ_clean_data %>% ggplot() + 
  geom_boxplot(aes(x = island_side, y = ammonia), na.rm = TRUE)

WQ_clean_data %>% ggplot() + 
  geom_jitter(aes(x = island_side, y = ammonia))
```



```{r}
WQ_clean_data %>% ggplot() + 
  geom_jitter(aes(x = site_type, y = salinity))
```

```{r}
WQ_clean_data %>% ggplot() + 
  geom_jitter(aes(x = site_type, y = water_temp))
```

```{r}
WQ_clean_data %>% ggplot() + 
  geom_jitter(aes(x = site_type, y = dissolved_oxygen))
```

```{r}
WQ_clean_data %>% ggplot() + 
  geom_jitter(aes(x = site_type, y = ammonia))
```

```{r}
WQ_clean_data %>% ggplot() +
  geom_jitter(aes(x = Month, y = water_temp, color = site_type))
```

```{r}
WQ_clean_data %>% ggplot() +
  geom_jitter(aes(x = Month, y = ph, color = site_type))

```

```{r}
WQ_clean_data %>% ggplot() +
  geom_jitter(aes(x = Month, y = dissolved_oxygen, color = site_type))
```


To Do:
1. Make Bay vs Ocean plots for each paremeter
2. Make a plot of site type for each parameter
3. geom_line for view hurricane Irma