---
title: "Hurricane Irma Analysis"
author: "William Norfolk"
date: "11/5/2019"
output: word_document
---

Based on the results of the exploratory and bivariate analyses a deeper dive into the specific trends of Hurricane Irma are warrented. 

First load the libraries.

```{r load libs}
library(readxl)
library(dplyr)
library(tidyverse)
library(forcats)
library(ggthemes)
library(plotly)
library(knitr)
library(naniar)
library(broom)
library(gridExtra)
library(zoo)
```


Next load the clean data and take a look.

```{r}
WQ_clean_data <- readRDS("../../data/processed_data/processeddata.rds")

glimpse(WQ_clean_data)
```

For this analysis we will subset the data to isolate the time frame close to Irma landfall. Landfall took place on September 10, 2017. We will selct data from 2017-2018 to view hurricane effects

```{r subset data}
year_2017 <- subset(WQ_clean_data, Year == "17")
year_2018 <- subset(WQ_clean_data, Year == "18")

#We also need to recode month for 2018 so it adds continiously.

year_2018 <- year_2018 %>% dplyr::mutate(Month = recode(Month,
                                  "01" = "13",
                                  "02" = "14",
                                  "03" = "15",
                                  "04" = "16",
                                  "05" = "17",
                                  "06" = "18",
                                  "07" = "19",
                                  "08" = "20",
                                  "09" = "21",
                                  "10" = "22",
                                  "11" = "23",
                                  "12" = "24"))

irma_subset <- rbind(year_2017, year_2018)

#Then convert Month back to numeric
irma_subset$Month <- as.numeric(as.character(irma_subset$Month))
```

Lets take a quick look at the parameters for this subset.


```{r}
irma_temp_plot <- ggplot(irma_subset, aes(x = Month, y = water_temp, color = island_side)) + geom_jitter(alpha = 0.5) + theme(legend.position = "bottom") + geom_vline(xintercept = 9.33, color = "black", linetype = "dotted", size = 1)

irma_temp_plot

mean_temp_irma <- aggregate(water_temp ~ Month + Year +island_side, irma_subset, mean)

irma_mean_temp_plot <- ggplot(mean_temp_irma, aes(x = Month, y = water_temp, color = island_side)) + geom_line(alpha = 1) + theme(legend.position = "bottom") + geom_vline(xintercept = 9.33, color = "black", linetype = "dotted", size = 1)

irma_mean_temp_plot + geom_point()
```

```{r}
irma_sal_plot <- ggplot(irma_subset, aes(x = Month, y = salinity, color = island_side)) + geom_jitter(alpha = 0.5) + theme(legend.position = "bottom") + geom_vline(xintercept = 9.33, color = "black", linetype = "dotted", size = 1)

irma_sal_plot

mean_sal_irma <- aggregate(salinity ~ Month + Year +island_side, irma_subset, mean)

irma_mean_sal_plot <- ggplot(mean_sal_irma, aes(x = Month, y = salinity, color = island_side)) + geom_line(alpha = 1) + theme(legend.position = "bottom") + geom_vline(xintercept = 9.33, color = "black", linetype = "dotted", size = 1)

irma_mean_sal_plot + geom_point()
```

```{r}
irma_amm_plot <- ggplot(irma_subset, aes(x = Month, y = ammonia, color = island_side)) + geom_jitter(alpha = 0.5) + theme(legend.position = "bottom") + geom_vline(xintercept = 9.33, color = "black", linetype = "dotted", size = 1)

irma_amm_plot

mean_amm_irma <- aggregate(ammonia ~ Month + Year +island_side, irma_subset, mean)

irma_mean_amm_plot <- ggplot(mean_amm_irma, aes(x = Month, y = ammonia, color = island_side)) + geom_line(alpha = 1) + theme(legend.position = "bottom") + geom_vline(xintercept = 9.33, color = "black", linetype = "dotted", size = 1)

irma_mean_amm_plot + geom_point()
```


```{r}
irma_do_plot <- ggplot(irma_subset, aes(x = Month, y = dissolved_oxygen, color = island_side)) + geom_jitter(alpha = 0.5) + theme(legend.position = "bottom") + geom_vline(xintercept = 9.33, color = "black", linetype = "dotted", size = 1)

irma_do_plot

mean_do_irma <- aggregate(dissolved_oxygen ~ Month + Year +island_side, irma_subset, mean)

irma_mean_do_plot <- ggplot(mean_do_irma, aes(x = Month, y = dissolved_oxygen, color = island_side)) + geom_line(alpha = 1) + theme(legend.position = "bottom") + geom_vline(xintercept = 9.33, color = "black", linetype = "dotted", size = 1)

irma_mean_do_plot + geom_point()
```

```{r}
irma_ph_plot <- ggplot(irma_subset, aes(x = Month, y = ph, color = island_side)) + geom_jitter(alpha = 0.5) + theme(legend.position = "bottom") + geom_vline(xintercept = 9.33, color = "black", linetype = "dotted", size = 1)

irma_ph_plot

mean_ph_irma <- aggregate(ph ~ Month + Year +island_side, irma_subset, mean)

irma_mean_ph_plot <- ggplot(mean_ph_irma, aes(x = Month, y = ph, color = island_side)) + geom_line(alpha = 1) + theme(legend.position = "bottom") + geom_vline(xintercept = 9.33, color = "black", linetype = "dotted", size = 1)

irma_mean_ph_plot + geom_point()
```









